RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2013-12-05 doc/bioinformatics2013/RVD2.tex

A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$, where $\mu_j^{(p)}$ is the $p$th from ythe posterior distribution. The acceptance probability is then
\begin{equation}
	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
\end{equation}

@@

The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. A matched normal cells were derived from lymphoblastoid cells from whole blood~\citep{}. 
% TODO: cite the original publication here.
% TODO: check matched cells line origin
% double checked by Yuting, didn't find the matched cells line origin
% TODO: mention tumor purity here

@@

% TODO: discuss alignment protocol

@@

@@

RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:XXX to chr7:XXX. Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants. The two heterozygous variants had a loss-of-heterozygosity in the tumor sample. 
%TODO: add start and stop location here

@@

The original research describing this sample used Strelka to identify mutations. They identified chr7:154760439 as variant, but did not call the other four variants. In particular Strelka missed the two LOH events.

>>>>

A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$, where $\mu_j^{(p)}$ is the $p$th from the posterior distribution. The acceptance probability is then
\begin{equation}
	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
\end{equation}

@@

The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. A matched normal cells were derived from lymphoblastoid cells from whole blood. Sequencing libraries were prepared according to the protocol described in the original technical report~\cite{}. The raw FASTQ read files were aligned to hg19 igenome~\cite{} using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}. We used samtools mpileup to generate pileup files using hg19 igenome~\cite{} as reference sequence~\citep{Navin:2010gu}. The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
% TODO:cite the technical report.
% TODO: check matched cells line origin
% double checked by Yuting, didn't find the matched cells line origin
% TODO: mention tumor purity here

@@

% TODO: discuss alignment protocol. Done.
% TODO: rvd2 parameters. Done.
% TODO: Pubmed for peer-reviewed paper on HCC1187 sample. Done

@@

%TODO: This section doesn't make sense to me. Lets talk about what needs to be done here.
Considering HCC1187 sample is diploid, we did two tests based on posterior density using RVD2. First, we compared the posterior density $\tilde\mu_j^{\Delta}$ to find out if $\tilde\mu_j$ in case sample was significantly higher than that in control sample and vice verse.Threshold $\tau$ was set at $1\times10^{-3}$ to avoid error brought by computer limitation; confidence level $1-\alpha$ was set at 0.95. We did $\chi^2$ test afterwards to further remove false callings.  % The difference is constant at 1.74*10^{-4}
\begin{equation}\label{eqn:bayes_test_positive}
	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
\end{equation}
\begin{equation}\label{eqn:bayes_test_negative}
	\Pr( \tilde{\mu}_j^{\text{control}} - \tilde{\mu}_j^{\text{case}} \geq \tau ) > 1-\alpha,
\end{equation}

Second, we identified the germline status and somatic status of HCC1187 sample by looking into posterior density of control sample and case sample independently. Basically, We computed the empirical posterior probability of $\tilde\mu_j$ across interval $[\tau_\text{l},\tau_\text{u}]$ and compare it with confidence level $1-\alpha$,
\begin{equation}\label{eqn:bayes_test_interval}
	\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha.
\end{equation}

Confidence level $1-\alpha$ was set at 0.8 due to our lower confidence on germline and somatic status.Posterior density interval for reference, heterzygous mutation and homozygous mutation were set as $[0.0,0.2],[0.2,0.8],[0.8.1]$, respectively.
% End of section that is confusing to me.

@@

RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774. Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants. The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.

@@

The original research describing this sample used Strelka to identify mutations in the same sample. They identified chr7:154760439 as variant, but did not call the other four variants. In particular strelka missed the two LOH events. % To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample. Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
